Supplementary material

### 2020-04-23 analysis/paper/supplementary.Rmd
- Supplementary material rmd - Added images and tables to supplementary

---
title: "The Gravettian-Solutrean transition in westernmost Iberia: new data from the sites of Vale Boi and Lapa do Picareiro"
author:
  - Joana Belmiro: 
      email: jnetobelmiro@gmail.com
      institute: [ICArEHB]
      correspondence: true
  - João Cascalheira:
      email: jmcascalheira@ualg.pt
      institute: [ICArEHB]
      correspondence: false
  - Nuno Bicho:
      email:
      institute: [ICArEHB]
  - Jonathan Haws:
    email:
    institute: [UL] [ICArEHB]
  - Ualg: ICArEHB, Universidade do Algarve, Portugal
  - UL: University of Louisville
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template2.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: bibliography.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  The present study aims to answer the question: what impact did the Heinrich event 2 have on the technological organization of human communities, at the onset of the Last Glacial Maximum, in south western Iberia? The impact of this event on the Gravettian-Solutrean transition has been previously suggested [@bradtmoller2012]. However, the existing models do not consider the Proto-Solutrean technocomplex as an individual phase for this transition [@cascalheiraandbicho2013].
  To address this question, this study analysed the lithic assemblages from Layers 5 and 4E of the site of Vale Boi (southern Portugal) and Levels U and T from Lapa do Picareiro (central Portugal). We aimed to understand the technological patterns and raw material exploitation during the Proto-Solutrean, and test the existing models with assemblages from recently excavated sites, while expanding the geographic range.
  The analysis followed a technological attributes approach. The retrieved data was fully analysed in R environment.
  Results show the existence of two discrete phases in each site. The first, with high frequency of quartz use for bladelet production, seems to reflect the presence, in both sites, of a Terminal Gravettian horizon, as defined by Almeida (2000). The second, with some significant differences between sites, attests the presence of Vale Comprido technology and lower quartz frequencies at Vale Boi, representing a Proto-Solutrean occupation; and the presence of a blade component in Lapa do Picareiro that, together with the respective absolute chronology, may attributed to a Proto-Solutrean or an Early Solutrean horizon.
  In general terms, this study allowed to confirm that the Terminal Gravettian and the Proto-Solutrean are discrete phases across the transiton, in concordance with the Three-Phase model presented by Zilhão [-@zilhao1997]. It further consolidates the expansion of similar techno-cultural patterns to southern Portugal, that may be explained by the expansion of social networks [@cascalheiraandbicho2013]. Associated with the dominance of different technological patterns and intensive use of quartz, we may understand these horizons as a moment of cultural reorganization, onset by environmental pressures.
keywords: |
  Upper Paleolithic; Stone tool analysis; Climate change; Portugal
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300,
  fig.ext = "tiff"
)
options(kableExtra.auto_format = FALSE)
```

```{r packages}

# Call packages
library(readr)
library(dplyr)
library(stringr)
library(tidyr)
library(knitr)
library(tab)
library(ggplot2)
library(FactoMineR)
library(factoextra)
library(RcmdrMisc)
library(IDPmisc)
library(forcats)
library(kableExtra)
library(float)
library(janitor)
library(ggpubr)
library(ggExtra)
library(ggsci)
library(ggalluvial)

@@

# Introduction

The trantion from the Gravettian to the Solutrean, impacted by adverse climatic conditions during the Heinrich Event 2 (HE2) and the onset of the Last Glacial Maximum (LGM), continues to be an essential topic to understand the Upper Paleolithic evolution and human adaptations to climatic changes at the end of the Pleistocene in Westernmost Europe.

There has been a significant amount of proposals suggesting a full synchronism between the onset of each Upper Paleolithic technocomplex and the occurrence of the most severe climate events [@cascalheiraandbicho2013; @bichoetal2017]. Bradtmöller et al. [-@bradtmoller2012] suggest a direct relationship between three of the HEs (4, 3 and 2) and the substitution of Neanderthal populations with the emergence of the Aurignacian, the appearance of the Gravettian, and the development of the Solutrean, respectively (Table \@ref(tab:climatetable)). Based on the theoretical framework of Panarchy [@holling2002], the authors propose the Repeated Replacement Model (RRM), where the HEs are understood as the primary climatic triggers for population turnover, through the breakdown of communication networks and cultural traditions, which were subsequently reorganized under different socio-cultural conditions [@bradtmoller2012].

```{r climatetable}
climate_table <- as_tibble(read_csv(here::here("analysis/data/raw_data/climate_table.csv")))
knitr::kable(climate_table,
             caption = 'Correspondence between Upper Paleolithic cultural phases (Aurignacian, Gravettian and Solutrean), Heinrich Events and D-O cycles, with respective chronology. Following dates and chronologies from Bradtmöller et al. (2012), Cascalheira and Bicho (2013), Sanchéz-Goñi and Harrison (2010).')
```

In this framework, the rapid climatic changes brought by the HE 2 were the trigger for the reorganization of human groups at the end of the Gravettian. Records indicate that during HEs, and with particular severity in the HE 2, western and southern Iberia must have experienced abrupt environmental changes, which altered the location, abundance and availability of resources for the communities of hunter-gatherers [@cascalheiraandbicho2013]. In some parts of Iberia and southwestern France these changes seem to be accompanied by significant techno-typological modifications, combining both elements from the previous and succeeding phases. For some authors, the temporal and spatial patterning of the elements that mark the transiton is, however, sufficiently significant to individualized different technocomplexes - such as the case of the Proto-Solutrean (@cascalheiraandbicho2013; @renard2011; @zilhaoetal1995; @zilhao1997).

Although an essential piece to understand how hunter-gatherer communities adapted, reinvented and abandoned their social and technological organization, most of what is known for the Proto-Solutrean is still geographically constricted and lacking good chronological markers [@cascalheiraandbicho2013], which hampers the understanding of the Proto-Solutrean solutions and their role in the Gravettian-Solutrean transition.

Result of the study of materials from old excavations by Manuel Heleno in Rio Maior (Portuguese Estremadura) in the 1980s, and new archaeological works in the same region, the Proto-Solutrean has thus been described as a transitional technocomplex from Gravettian to Solutrean technologies, through a process of local development, and synchronous to all of southwestern Europe [@zilhaoetal1999].

In the westermost edge of Iberia this technocomplex is best known from open-air and cave sites located in the Portuguese Estremadura (Figure \@ref(fig:protomap)).

```{r protomap, fig.cap="Location of Proto-Solutrean sites in Portugal or with Proto-Solutrean finds, including the two sites analysed in this study. Legend: 1-Buraca Escura; 2-Alecrim e Lagar Velho; 3-Caldeirão; 4-Lapa do Anecrial; 5-Lapa do Picareiro; 6-Terra do José Pereira; 7-Terra do Manuel; 8-Vales de Senhora da Luz; 9-Vale Comprido - encosta; 10-CPM III; 11-Gato Preto; 12-Gruta das Salemas; 13-Vale Boi"}
knitr::include_graphics(here::here("analysis/figures/map_proto_solutrean.png"))
```

Here, the Proto-Solutrean has been integrated into the Gravettian-Solutrean transition as part of a sequence of techno-typological mofications that could have occured across two or three different stages [@zilhao1994; @zilhao1997; @zilhaoetal1999].

The Two-stage model starts with a Final Gravettian phase, characterized by a moderate use of quartz (~15%), production of truncated backed bladelets and proto-magdalenian retouched blades [@zilhao1997]. The second phase is the Proto-Solutrean, frequently characterized by a high percentage of quartz use (~30%), rare presence of backed bladelets or, when retouch is present, being marginal, and the production of Vale Comprido points, the technocomplex’s index fossil [@almeida2000; @zilhao1997; @zilhaoetal1995]. For the production of these tools, the Proto-Solutrean shows three different operative sequences: one for the production of Vale Comprido points, through the removal of elongated blanks with convergent profiles and thick platforms; another for the production of blades, of Gravettian tradition; and another for the production of bladelets, probably obtained through the exploitation of thick endscapers or carinated elements [@zilhao1997].

Vale Comprido points are described as robust pieces, often characterized by convergent shapes, triangular cross-sections and plain platforms, often having a high elongation ratio, although not necessarily falling into the blade category. The Vale Comprido points have been used to describe the technological transition between the Proto-Solutrean and the Solutrean. In this case, these points are seen as an element of discontinuity with the previous technocomplex, where the organic points armed with microliths from the Gravettian are replaced by lithic points, with enough similarities to the *pointes à face plane* of the Middle Solutrean to be understood as a technological development [@zilhao2013].

The Three-stage model maintains the Final Gravettian phase and its defining characteristics but subdivides the following Proto-Solutrean stage in two. As such, there is an intermediate stage characterized by the intensive use of quartz (~30%), which corresponds to Laugerie-Haute’s Aurignatian V (also known as Terminal Gravettian). The third stage is the Proto-Solutrean, where quartz use diminishes. In this model, the Vale Comprido points and associated reduction sequence may appear in either of the last two stages [@zilhao1997]. Alternatively, in the Two-phase model, the Terminal Gravettian is understood as a functional facies, related to specialized occupations and production activities within the technocomplex.

Regardless of what model is accepted, the Proto-Solutrean, stands out for its technological innovations. These are reflected in the manufacture of Vale Comprido points and a complete operative sequence for their production, but marked by a high degree of technological variability in their lithic assemblages [@almeida2000; @zilhao1997; @zilhaoetal1999].

This variability has been interpreted as a technological race in response to the environmental modifications taking course during the HE 2 [@cascalheiraandbicho2013]. The authors suggest that, in order to correspond to the external pressures, there may have been the need to diversify the economic strategies in use until that moment. The high exploitation of quartz, for example, formerly a secondary raw material, and the use of similar reduction strategies between quartz and chert, might represent one such economic response. The same possibly applies to the use of unprecedented raw materials, as is the case of the Proto-Solutrean levels in Vale Boi, where there is the presence of previously unknown raw materials, used mainly for the manufacture of Vale Comprido points [@belmiro2017; @marreiros2009]. Likewise, alterations recorded in territoriality patterns for the Proto-Solutrean can be interpreted has extensive regional networks and may be related to modifications in the landscape that occurred during the HE 2 [@cascalheiraandbicho2013; @cascalheira2019].

However, despite the existence of rather comprehensive literature about the technocomplex in the Portuguese Estremadura, which has allowed for a better understanding of the impacts of HE 2 in hunter-gatherer communities and the emergence of the Solutrean, there is still the need for further studies, in order to fill existing gaps. One of these gaps is the concentration of data regarding this technocomplex in the Estremadura, whereas the Proto-Solutrean is still relatively unknown in other places throughout southwestern Europe. In the case of Portugal, only one other site in the south (Figure \@ref(fig:protomap)) has so far revealed the presence of this techno-complex. Another issue is the small amount of absolute datings for Proto-Solutrean contexts, which limits the chronological definition of the technocomplex, and thus the testing of which transition model (if any) applies best [@cascalheiraandbicho2013].

Addressing these issues, through the study of lithic assemblages from other areas in southwestern Europe, from sites with good stratigraphic preservation which allow absolute dating and accurate spatial tracking, will undoubtedly help further understand the transition, its patterns, stage transitions and possible regional variations.

By understanding the need to better comprehend the Proto-Solutrean and how abrupt climate change impacted these communities, this study's presents the results of the analysis of stone tool assemblages coming from levels attributed to the Gravettian-Solutrean transition timeframe, from two recently excavated sites - Vale Boi (southern Portugal) and Lapa do Picareiro (central Portugal).

Specificallt, by combining techno-typological information with stratigraphy and absolute chronology results, we present new data on: 1) the technological organization of levels 5/4E of Vale Boi, and levels U/T from Lapa do Picareiro; 2) a stratigraphical and chronological separation between Terminal Gravettian horizons and Proto-Solutrean occupations at both sites, revealing the application of the same general set of pattern to geographic areas outside of the Estremadura; 3) and, consequently, the validity of the three-stage model as the best suited for the currently available data.

# Materials and methods

The materials used for this analysis include all lithic artifacts from Vale Boi's layers 5 and 4E of the Terrace area and levels U and T from Lapa do Picareiro (see below).

For lithic analysis we adopted an attribute-based methodology, which aims to describe morphological and metrical attributes of technological classes. The attributes analyzed followed those present in specialized literature, such as @brezillon1968 and @tixier1980, paired with apprcascalheira2010,hes used in o)ther Upper Paleolithic lithic attribute analysis (see e.g. @bicho1992, @zilhao1997, @cascalheira2010, @almeida2000).

Artifacts and their techno-morphological attributes were recorded using two different databases, thus dividing the analysis in two phases: basic and complete attribute-based analysis.

The basic analysis was applied to all of Vale Boi’s quartz artifacts (layers 5 and 4E), since all lithics from the same assemblage in other raw-materials (*e.g.* chert and greywacke) had been previously analysed with the same database  [see @belmiro2018]. This database consists of a reduced set of variables which allow for the general classification of metric and morphologic characteristics of the artifacts, as well as the identification of raw material preference and preferential use.

The complex analysis was applied to Vale Boi cores, debitage products and retouched pieces. Only pieces which were complete were chosen for this stage for two main reasons: 1) the collection is well preserved, with high percentages of complete cores and blanks (~80%), guaranteeing that the sample is representative; 2) only complete pieces gather all the necessary attributes for a thorough analysis. The same database was applied to the whole assemblage of Lapa do Picareiro, given its small sample size.

Data collection was done through E4, a software developed by Dibble and McPherron (2003). The complete Data Dictionary with all variables, attributes, values and description, with reference to the consulted literature is available in our research compendium at https://doi.org/xxxx.

After data collection, the databases were imported into R environment, where the information was processed through the creation of descriptive statistical analysis and writing of this article. We include the entire R code used for all the analysis and visualizations contained in this paper is in our online research compendium. To produce those files, we followed the procedures described by Marwick et al. (XXXX) for the creation of research compendiums to enhance the reproducibility of research. To enable maximum re-use, our code is released under the MIT license, our data as CC-0, and our figures as CC-BY (for more information, see Marwick XXXX).

@@

# Vale Boi

## Stratigraphy and chronology

Vale Boi is an open-air site and rockshelter, located on the western coast of Algarve (Portugal), near a small homonymous village, within the municipality of Vila do Bispo (Figure \@ref(fig:protomap)). The site is situated in a small valley that runs south to the Atlantic coast, about 2 km distance, relatively open, with a natural boundary to the east and bordered by a limestone hill through all its extension. This hill is marked, at specific points, by limestone exposures that form rock shelters with faces facing west or southwest (Figure \@ref(fig:vbphoto))[@bicho2003; @cascalheira2010; @cascalheiraetal2008].

```{r vbphoto, fig.cap="Vale Boi. Topographic plan, schematic profile and general view, with location of the excavation loci. After Cascalheira et al. (2017)."}

The site extends for more than 10 000 $m^2$ on the slope of this valley, which is marked by a series of steps that run parallel to the river, possibly the result of Middle Pleistocene fluviatile erosion [@bicho2003].

The geological context of Vale Boi is marked by heterogeneity. In the north, there are schist and greywacke formations from the Carboniferous. In the south, Triassic and Jurassic dolomite and limestone formations, which are gradually covered by Holocene dunes further into the coastal area, until near St. Vincent’s cape where they appear uncovered once again, along with small occurrences of chert [@verissimo2004].

Vale Boi shows a variety of human occupations, distributed across the three main areas, which have been interpreted differently in terms of functionality: Slope, Shelter, and Terrace.

Although excavated since 2000, in 2012, a new 8 $m^2$ area was open in the Terrace (rows H and I), to understand the stratigraphic sequence in more detail and assess the existence of older cultural horizons, attributable to an early Upper Paleolithic occupation.

Since 2012, six layers have been identified (Table \@ref(tab:vbstrat)). In some of these layers, lateral sediment variations were identified, which were coded in the field through the concatenation of a letter to the layer number (*e.g.* 4E). In many of these cases, the isolation of this vertical variation did not show any patterns in terms of spatial concentration of materials, although in others, like the 4E facies, the subdivision of the layer correlated with the spatial distribution of Vale Comprido technology.

In the new area, layers 1 and 2 represent Holocene levels, with the first layer being possibly disturbed by agricultural processes. Layer 2, with a thickness of 25-30 cm shows a Neolithic occupation.

Layer 3 has a silt and clay matrix sediment, with some inclusions and showing interruptions of limestone clast depositional episodes, although there is the constant presence of fauna and lithic artifacts. As mentioned above, the different material and cultural characteristics within the same geologic package led to the subdivision of the layer in layer 3A, attributed to an Epipaleolithic occupation, and 3B, assigned to the Solutrean.

Layer 4 is very similar to layer 3. However, it is separated from it by a gravel level. Similarly to layer 4, these layers have been subdivided regarding different degrees of sediment compaction and/or concentration of organic materials, showing two differing cultural horizons: Solutrean and Proto-Solutrean (limited to layer 4E).

Layer 5 has a dark coloration and a silt and clay matrix characterized by an intense presence of organic elements, frequently calcinated. There is the presence of a Proto-Solutrean horizon within the top levels, although occupation intensity seems to diminish with depth.

Finally, layer 6 is very similar to the previous layer, although it shows the presence of a larger quantity of small and medium-sized limestone clasts. A Gravettian horizon has been attributed to this layer, but the analysis of the materials is currently in progress.

Regarding layers 4E and 5, although they are indeed layers with different sedimentary packages, they show similar technological and archaeological patterns, which led to the conclusion they were the same cultural horizon (the opposite, one single layer with different cultural horizons is equally present in the Terrace, as seen with the epipaleolithic and Solutrean occupations of layer 3).

These layers, 4E and 5, are vertically contiguous, separated by a relatively flat surface covered by large limestone blocks. Unlike what was expected, however, these levels did not reveal a Gravettian assemblage with characteristics similar to previous years [@marreiros2009], but rather materials with patterns similar to those expected in a Proto-Solutrean assemblage (the most noticeable of all being the Vale Comprido points and blanks). The context was dated using a charcoal sample to c. 24.7-25.3 ka cal BP, and a shell sample to c. 23.7-24.1 ka cal BP (Table \@ref(tab:protodates)).

The bottom spits of layer 5, on the other hand, show a relatively dense concentration of *Littorina littoreia*, an unprecedented shell species at the site, and often associated with colder waters, having a high freezing tolerance [@murphy1979; @clarke2000]. This context was dated through a shell sample, showing dates ranging between the 26-26.3 ka cal BP, which corresponds to the first half of the HE 2 [@sanchez-goni2010].

```{r c14VB, echo = FALSE, fig.cap = "Vale Boi. Radiocarbon results for layers 5 and 4. The results presented are modeled dates based on a Bayesian model (Ramsey et al. 2009) using all dates available for the stratigraphic sequence of the Terrace area. All dates were calibrated using the IntCal13 and Marine13 curves (Reimer et al. 2013), using a regional Delta value of 265+/- 107 in the case of marine shells. The Oxcal script bayesian model is provided as online supplementary materials.", fig.align= 'center', out.width = '60%', fig.pos="H"}

knitr::include_graphics(here::here("/analysis/figures/c14_VB.png"))

One of the most noticeable features about the vertical distribution of pieces in the Terrace (row H and I; Figure \@ref(fig:spatialvbfig)) is the vertical constraint in the distribution of dolerite pieces, mostly concentrated on Layer 4E and upper levels of Layer 5, coinciding also with the higher concentration of lithic materials in these layers. Likewise, the presence of Vale Comprido points is also associated with Layer 4E and the top levels of Layer 5 [@belmiro2017], which seems to occur around 25 ka cal BP (Figure \@ref(fig:c14VB)). The WK-44416 date seems to show that, at least from around c. 26 ka cal BP, the site was occupied even if with lower intensity, a date that falls onto the Final Gravettian/Proto-Solutrean or Final Gravettian/Terminal Gravettian transition in the traditional models [@zilhao1997].

The radiocarbon results represented on Figure \@ref(fig:c14VB) seem to show that there are three distinct chronological moments within layer 5. Given the horizontal distribution of the dates from upper layer 5, it may seem, however, that the dates which fall onto the c. 24 ka cal BP are intrusions from the upper layer.

## Lithic assemblages

Given the possible existence of two phases within the Proto-Solutrean occupations (Three-phase model), we calculated the percentages of the three most important raw materials by cubic meter of excavated sediment (Figure \@ref(fig:rmdispersion)) and plotted them by depth. There seem to be significant differences in raw material distribution over time. From around 24.1 m of depth upwards, there is an important shift in quartz and chert frequencies, the latter increasing more than 10%, and quartz dropping from c. 50% to nearly 30%. Greywacke frequencies follow those of quartz.

This shift seems to be associated with other significant changes in the Terrace sequence, such as an increase in the amount of lithic materials in top of Layer 5 and Layer 4E, but also the appearance of Vale Comprido technology. Additionally, these two moments are stratigraphically correlated with two different chronological horizons, the first, dated to c. 26 ka cal BP, at c. 23.9 m depth, and associated with higher frequencies of quartz, and the second, dated to c. 24.7 ka cal BP, at around 24.1 m depth, associated with higher frequencies of chert and a reduction in quartz presence.

```{r rmdispersion, fig.cap = "Vale Boi. Raw material discard rates over time. Each point is an excavation unit (spit). The lines are locally weighted regression lines (span ¼ 0.4) to aid in visualising the trend of increased discard in the upper part of the deposit.", echo = FALSE, message=FALSE, warning=FALSE}

# Calulate spit volumes
cols_to_concat <- c("Level", "Spit")

buckets_by_spit <- context %>%
  filter(Level %in% c("4E", "5", "5B", "5C") & Code == "BUCKET" & !Spit %in% c("limp_corte", "profile", " ", NA)) %>%
  unite_(col='depth', cols_to_concat, sep=".", remove=FALSE)

buckets_by_spit$depth <- str_replace(buckets_by_spit$depth, "B", "")
buckets_by_spit$depth <- str_replace(buckets_by_spit$depth, "C", "")

spit_vol <- buckets_by_spit %>%
  dplyr::group_by(depth) %>%
  dplyr::count(Code) %>%
  mutate(spit_vol = (n*10)/1000)

# Calculate average depth of spits

spit_depths <- field_data %>%
  filter(Level %in% c("4E", "5", "5B", "5C") & !Spit %in% c("limp_corte", "LIMP_CORTE", "profile", " ", NA, "0")) %>%
  unite_(col='depth', cols_to_concat, sep=".", remove=FALSE)

spit_depths$depth <- str_replace(spit_depths$depth, "B", "")
spit_depths$depth <- str_replace(spit_depths$depth, "C", "")

spit_aver_depths <- spit_depths %>%
  dplyr::group_by(depth) %>%
  dplyr::summarise(avg = mean(Z)) %>%
  mutate_if(is.numeric, format, 3)

## Data classes per spit

cols_to_concat <- c("Level.y", "Spit")

lithics_by_spit <- dataset %>%
  filter(Level.y %in% c("4E", "5", "5B", "5C") & !Spit %in% c("limp_corte", "profile", " ", NA)) %>%
  select(id, Z, Spit, Level.y, Code, Class, ChipQuantity, RetouchedPieceTypology, PieceCompleteness, RawMaterial) %>%
  unite_(col='depth', cols_to_concat, sep=".", remove=FALSE)

lithics_by_spit$depth <- str_replace(lithics_by_spit$depth, "B", "")
lithics_by_spit$depth <- str_replace(lithics_by_spit$depth, "C", "")

#Debitage by spit

cols_to_concat <- c("Level.y", "Spit")

rm_by_spit <- dataset %>%
  filter(Level.y %in% c("4E", "5", "5B", "5C") & !Spit %in% c("limp_corte", "profile", " ", NA)) %>%
  select(id, Z, Spit, Level.y, Code, RawMaterial, QuartzQuality, Class) %>%
  unite_(col='depth', cols_to_concat, sep=".", remove=FALSE) %>% 
  filter(Class %in% c("Blank","Core","RetouchedPiece"))

rm_by_spit$depth <- str_replace(rm_by_spit$depth, "B", "")
rm_by_spit$depth <- str_replace(rm_by_spit$depth, "C", "")

rm_by_spit <- rm_by_spit %>% 
  dplyr::group_by(depth) %>%
  dplyr::count(RawMaterial) %>%
  mutate(freq = n / sum(n)*100)

quartz_by_spit <- rm_by_spit %>%
  filter(RawMaterial == "Quartz")

chert_by_spit <- rm_by_spit %>% 
  filter(RawMaterial == "Chert")

grey_by_spit <- rm_by_spit %>% 
  filter(RawMaterial == "Greywacke")

quartz_by_spit$spit_vol <- spit_vol$spit_vol[match(quartz_by_spit$depth, spit_vol$depth)]
quartz_by_spit <- quartz_by_spit %>%
  mutate(artifacts_by_cubic_meter = n/spit_vol)
quartz_by_spit$spit_aver_depth <- spit_aver_depths$avg[match(quartz_by_spit$depth, spit_aver_depths$depth)]
quartz_by_spit$spit_aver_depth <- as.numeric(as.character(quartz_by_spit$spit_aver_depth))

chert_by_spit$spit_vol <- spit_vol$spit_vol[match(chert_by_spit$depth, spit_vol$depth)]
chert_by_spit <- chert_by_spit %>%
  mutate(artifacts_by_cubic_meter = n/spit_vol)
chert_by_spit$spit_aver_depth <- spit_aver_depths$avg[match(chert_by_spit$depth, spit_aver_depths$depth)]
chert_by_spit$spit_aver_depth <- as.numeric(as.character(chert_by_spit$spit_aver_depth))

grey_by_spit$spit_vol <- spit_vol$spit_vol[match(grey_by_spit$depth, spit_vol$depth)]
grey_by_spit <- grey_by_spit %>%
  mutate(artifacts_by_cubic_meter = n/spit_vol)
grey_by_spit$spit_aver_depth <- spit_aver_depths$avg[match(grey_by_spit$depth, spit_aver_depths$depth)]
grey_by_spit$spit_aver_depth <- as.numeric(as.character(grey_by_spit$spit_aver_depth))

rm_by_cubic_meter <- bind_rows(quartz_by_spit, chert_by_spit, grey_by_spit)

rm_by_cubic_meter <- filter(rm_by_cubic_meter, spit_vol > 0.10) # removes low volumetric levels

# Plot

rm_cubic_meter <- ggplot(rm_by_cubic_meter, (aes(spit_aver_depth, freq, colour = RawMaterial))) +
  geom_point() +
  stat_smooth(span = 0.5, se = FALSE) +
  xlab("Depth (m)") +
  ylab("Artefact relative frequency by cubic meter of sediment (log10)") +
  theme(axis.text=element_text(size=12),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        legend.text = element_text(size= 12),
        legend.title = element_text(size = 14)) +
  scale_y_log10()+
  coord_flip() +
  scale_color_jco()

rm_cubic_meter
```

Given the chronological data, density of artifacts, and raw material preference patterns, it was decided that for this study the materials would be subdivided in two analytical units, to better understand any possible technological differences: **Lower 5**, including all artifacts with Z values under 24.1; **Upper 5/4E**, including all artifacts with Z values of 24.1 and above. The value 24.1 is arbitrary and was chosen for the reasons aforementioned: it seems to be where the separation between higher and lower densities of lithic materials occurs and, also, where the inversion of quartz and chert frequencies happens.

A total of 26,703 pieces were analysed for both groups, 11,094 from Lower 5 group (Table \@ref(tab:general1)) and 15,609 for the Upper 5/4E (Table \@ref(tab:general2)). Most of these are chips (70.62% from Lower 5 and 65.68% from Upper 5/4E), followed by shatters, which make up c. 21% and c. 24% of the groups, respectively. These extremely high numbers for debris are mostly the result of quartz use, which can be explained not only by on-site knapping of this raw material but mostly by its breakage patterns (especially when coarser), which typically produces more waste than in other raw materials.

For Lower 5, cores and debitage products represent a cumulative frequency of c. 8%. For Upper 5/4E, debitage products represent nearly 10% of the group's assemblage. Complete blanks are the most represented class for both groups, with 1,474 identified pieces, 4.59% for Lower 5 and 6.18% for Upper 5/4E, followed by blank fragments, with an absolute count of 248 for Lower 5 and 336 for Upper 5/4E. Cores are also relatively frequent within these assemblages, with an absolute count of 123 cores, 46 for Lower 5 and 77 for Upper 5/4E, while core fragments appear in much smaller numbers (n=11 and n=14, respectively).

During the analysis, 167 retouched pieces were identified, 55 on Lower 5, representing 0.5% of the group, and 112 on Upper 5/4E, with a frequency of 0.72%. Retouched piece fragments appear in much smaller numbers (n=7 for Lower 5 and n=16 for Upper 5/4E).

One of the most relevant raw materials in both of Vale Boi's analytical units used for this study is quartz (encompassing rock crystal, fine, medium and coarse qualities), which represents c. 43% of the total assemblage for the Upper 5/4E group, and c. 51% for Lower 5. Chert, on the other hand, represents c. 46% of the Upper 5/4E phase and 38% of the Lower 5. Finally, greywacke, dolerite, chalcedony, and other raw materials that due to their low frequencies were collapsed in the category “Other”, together represent c. 11% of both assemblages (Figure \@ref(fig:rmvb)).

```{r rmvb, fig.cap="Vale Boi. Frequencies of raw materials by phase. Chips and shatters not included.", echo=FALSE}

dfRM_vb <- dataset %>% 
  filter(Class!="Chip") %>% 
  filter(Class!="Shatter") %>% 
  drop_na(Phase)

dfRM_vb %>%
  mutate(RawMaterial = factor(RawMaterial, levels = c("Quartz","Chert","Greywacke","Dolerite","Chalcedony","Other"))) %>%
  dplyr::group_by(Phase, RawMaterial) %>% 
  dplyr::summarise(n = n()) %>%
  mutate(percent = n/sum(n)) %>% 
  ggplot(aes(x=Phase, y = percent, fill = RawMaterial)) +
  geom_bar(stat = "identity", width = 0.5) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_minimal() + #type of theme
  labs(x = "Phases", y = NULL) +# change labels
  guides(fill=guide_legend(title=NULL)) + # remove legend title
  geom_text(aes(label=ifelse(percent >= 0.07, paste0(sprintf("%.0f", percent*100),"%"),"")),
                position=position_stack(vjust=0.5), colour="white") +
  theme(axis.title.x = element_text(size = 10, vjust = -1)) +
  scale_fill_jco()

The number of chips and shatter in quartz are extremely high, both in its representativity within the raw material, with c. 94% for both Lower 5 and Upper 5/4E, and when compared to the relative frequencies of other raw materials.

Despite the low frequencies presented on the table, there is a large number of quartz blanks (n=281 for Lower 5 and n=438 for Upper 5/4E) compared to the rest of the classes, as well as cores (n=18 for Lower 5 and n=32 for Upper 5/4E) and retouched pieces (n=15 for Lower 5 and n=34 for Upper 5/4E).

Chert shows high frequencies of blanks (15.8% for Lower 5 and 19.6% for Upper 5/4E) when compared to every other class, excluding chips. The latter represents more than 50% of the chert assemblages, with an absolute frequency of 649 chips for Lower 5 and 1,158 for Upper 5/4E.

Although chert shows smaller absolute numbers for blanks (n=171 for Lower 5 and n=407 for Upper 5/4E) and blank fragments (n=90 for Lower 5 and n=155 for Upper 5/4E) compared to quartz, there is a large quantity of cores representing 2.1% of total chert in Lower 5 and 2% in Upper 5/4E. However,there is a noticeable difference in the ratio between chert blanks and quartz blanks in Lower 5 and Upper 5/4E, with the latter showing a higher frequency of chert blanks compared to Upper 5/4E.

Regarding retouched pieces, these represent more than 3% of chert for both phases, a number far higher than those in other raw materials. This shows that chert, although less present than quartz in blank absolute count (although barely in Upper 5/4E), might have been preferentially used for formal tool production.

>>>>

---
title: "Supplementary material"
output: 
  pdf_document: 
    latex_engine: xelatex
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE
  )
```

```{r callpackages, include=FALSE}

# Call packages
library(readr)
library(dplyr)
library(stringr)
library(tidyr)
library(knitr)
library(tab)
library(ggplot2)
library(FactoMineR)
library(factoextra)
library(RcmdrMisc)
library(IDPmisc)
library(forcats)
library(kableExtra)
library(float)
library(janitor)
library(ggpubr)
library(ggExtra)
library(ggsci)
library(ggalluvial)
```

```{r datadictionary, echo=FALSE}
data_dictionary <- read_csv(here::here("analysis/data/raw_data/data_dictionary.csv"))
knitr::kable(head
(data_dictionary [1:4],46),longtable = TRUE, booktabs=TRUE, caption = "Data Dictionary with variables considered in the attributes analysis, including measurement units, allowed vallued, definitions and/or references.") %>% 
  landscape() %>% 
  kable_styling(latex_options = "repeat_header", font_size = 9) %>% 
  column_spec(1, width = "2cm") %>% 
  column_spec(2, width = "3cm") %>%
  column_spec(3, width = "6cm") %>% 
  column_spec(4, width = "9cm")

@@

```{r metric, fig.cap="Registered measurements scheme for debitage products and cores. Legend: 1- Maximum length; 2- Distal width; 3- Mesial width; 4- Maximum width; 5- Proximal width; 6- Mesial thickness; 7- Platform thickness; 8- Platform width.", out.width = '60%', fig.align='center', echo=FALSE}

knitr::include_graphics(here::here("analysis/figures/Metrics-01.png"))

```{r vbmap, fig.cap="Vale Boi. Location.",fig.align='center', out.width = '60%', echo=FALSE}

knitr::include_graphics(here::here("analysis/figures/vale_boi_map.png"))

```{r vbphoto, fig.cap="Vale Boi. Topographic plan, schematic profile and general view, with location of the excavation loci. After Cascalheira et al. (2017).", fig.align='center', out.width = '60%', echo=FALSE}

```{r layertable, include=FALSE}

layers_table <- read_csv(here::here("analysis/data/raw_data/VBlayers_table.csv"))
layers_table <- layers_table %>%
  na.omit()

```{r vbstrat, echo=FALSE}

knitr::kable(head(layers_table [1:2], 6),  booktabs = TRUE, caption = "Vale Boi. Terrace identified layers with sediment description.") %>% 
  column_spec(1, bold = T) %>% 
  kable_styling(font_size = 9) %>% 
  column_spec(2, width = "12cm") %>%
  column_spec(1, width = "2cm") %>% 
  kable_styling(latex_options = "hold_position")

```{r spatialvbfig, fig.cap="East profile stratigraphy in the Terrace area (above) and distribution of all tridimensionally coordenated lithics in the new excavation area until 2017 (under). Lithis from layers 4E and 5 are coloured in black. Blue crosses represent dolerite artefacts and stars mark the provenance of each radiocarbon date: A - WK-42830; B - WK-42831; C - WK-44416.", fig.align='center',out.width = '70%', echo=FALSE, fig.pos="H"}

knitr::include_graphics(here::here("analysis/figures/SpatialanalysisVB.png"))

@@
